---
title: "Practical"
author: "Remy Beugnon"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Day 1

```{r cars}
set.seed(123)

n = 50
y = rnorm(n, 1, 2)

y 

hist(y)

```

### Likelihood function

#### Gausian for given mean and sd

```{r}
curve(dnorm(x, mean = 0, sd = 1), 
      from = -4, to = 4)
```

#### Estimation of maximm likelihood
```{r}
i = 1 

L = dnorm(y[i], mean = 0, sd = 1)

L 

```

For all data
```{r}
L = dnorm(y, mean = 0, sd = 1)

L
```

### Maximization

#### Product of L 

```{r}
prod(L)
```

Log transformation to work on bigger number and avoid computation limitations.

```{r}
sum(log(L))
```

```{r}
curve(dnorm(x, mean = 0, sd = 1), 
      from = -4, to = 4)
for(i in 1:n){ lines(c(y[i], y[i]), c(0, L[i])) }

curve(dnorm(x, mean = 1, sd = 2), 
      from = -4, to = 4, col = 'red')
for(i in 1:n){ lines(c(y[i], y[i]), c(0, dnorm(y[i], mean = 1, sd = 2)), col = 'red')}
```

#### Log likelihood function 

```{r}
nll.function = function(y, par){
  LL = dnorm(x = y, mean = par[1], sd = par[2], log = T)
  NLL = - sum(LL)
  return(NLL)
}

nll.function(y, par = c(1, 2))
```

#### Look for optimum

```{r}
optim(par = c(1,1), 
      fn = nll.function,
      y = y)
```
### Model in Stan

Package

```{r, message=F, warning=F}
library(rstan)
library(coda)
```

Data

```{r}
n = 100

a = 1
b = 2
sigma = .5

x = runif(n, 0, 1)
y = a + b * x + rnorm(n, 0, sd = sigma)

df = data.frame(
  x = x, 
  y = y
)

plot(df)
```

Stan code

```{r}
stan_code = '
data{
  int n; // comments
  real x[n];
  real y[n];
}

parameters{
  real a;              //intercept
  real b;              //slope
  real<lower=0> sigma;    //error
}

model{
  // Priors
  real mu[n];
  a ~ normal(0,10);
  b ~ normal(0,10);
  sigma ~ normal(0,10);
  // Likelihood
  for(i in 1:n){
    mu[i] = a + b * x[i];
    y[i] ~ normal(mu[i], sigma); 
  }
}
'

stan_model = stan_model(model_code = stan_code)

data = list(n = nrow(df),
            x = df$x,
            y = df$y)

fit = sampling(stan_model, 
               data = data)
```

```{r}
print(fit)
```

### Run in parallel the chains

```{r}
rstan_options(auto_write = T)
options(mc.cores = 4)
```


```{r}
fit = sampling(stan_model, 
               data = data)

summary(fit)
```

```{r}
plot(fit)

plot(fit, pars = c('a', 'b'))
```

```{r}
stan_trace(fit)
```


```{r}
plot(As.mcmc.list(fit))

plot(As.mcmc.list(fit)[, 1:2])
```

```{r}
pairs(fit, 
      pars = c('a', 'b', 'sigma'))

```

```{r}
library(BayesianTools)

correlationPlot(as.matrix(fit))
correlationPlot(as.matrix(fit)[,1:3])
```


# Day 2

```{r}
# Model 
stan_code = '
data{
  int n; // comments
  real x[n];
  real y[n];
}

parameters{
  real a;              //intercept
  real b;              //slope
  real c;              //slope x2
  real<lower=0> sigma; //error
}

model{
  // Priors
  real mu[n];
  a ~ normal(0,10);
  b ~ normal(0,10);
  c ~ normal(0,10);
  sigma ~ normal(0,10);
  // Likelihood
  for(i in 1:n){
    mu[i] = a + b * x[i] + c * x[i] * x[i];
    y[i] ~ normal(mu[i], sigma); 
  }
}
'

stan_model = stan_model(model_code = stan_code)

data = list(n = nrow(df),
            x = df$x,
            y = df$y)

fit = sampling(stan_model, 
               data = data)


print(fit)
library(BayesianTools)
correlationPlot(as.matrix(fit)[,1:3])
```

Using vectors for the paameters to estimate

```{r}
# Model 
stan_code = '
data{
  int n; // comments
  vector[n] x;
  vector[n] y;
}

parameters{
  vector[3] parm;      //parameters
  real<lower=0> sigma; //error
}

model{
  // Priors
  real mu[n];
  parm ~ normal(0,10);
  sigma ~ normal(0,10);
  // Likelihood
  for(i in 1:n){
    mu[i] = parm[1] + parm[2] * x[i] + parm[3] * x[i] * x[i];
    y[i] ~ normal(mu[i], sigma); 
  }
}
'
```




```{r}
# Model 
stan_code = '
data{
  int n;
  vector[n] x;
  vector[n] y;
}

parameters{
  vector[3] parm;      //intercept
  real<lower=0> sigma; //error
}

model{
  // Priors
  vector[n] mu;
  parm ~ normal(0,10);
  sigma ~ normal(0,10);
  // Likelihood
    mu = parm[1] + parm[2] * x + parm[3] * x .* x;
    y ~ normal(mu, sigma); 
}
'
```

#### Parameters of the sampler

```{r}
fit = sampling(stan_model, 
               data = data,
               iter = 2000,   # sampling size
               warmup = 1000, # size of the warmup phase
               chains = 4     # number of chaines
              )
```


## 2.2 

### Effect of sample size

```{r}
stan_code = '
data{
  int n; // comments
  real x[n];
  real y[n];
}

parameters{
  real a;              //intercept
  real b;              //slope
  real<lower=0> sigma; //error
}

model{
  // Priors
  real mu[n];
  a ~ normal(0,10);
  b ~ normal(0,10);
  sigma ~ normal(0,10);
  // Likelihood
  for(i in 1:n){
    mu[i] = a + b * x[i];
    y[i] ~ normal(mu[i], sigma); 
  }
}
'

stan_model = stan_model(model_code = stan_code)


df.1 = df[sample(1:100, 10),]
data.1 = list(n = nrow(df.1),
            x = df.1$x,
            y = df.1$y)
fit.1 = sampling(stan_model, 
               data = data.1)
plot(fit.1)

df.2 = df[sample(1:100, 50),]
data.2 = list(n = nrow(df.2),
            x = df.2$x,
            y = df.2$y)
fit.2 = sampling(stan_model, 
               data = data.2)
plot(fit.2)

df.3 = df
data.3 = list(n = nrow(df.3),
            x = df.3$x,
            y = df.3$y)
fit.3 = sampling(stan_model, 
               data = data.3)
plot(fit.3)
```

## Effect of prior information

```{r, echo=F}
stan_code.1 = '
data{
  int n; // comments
  real x[n];
  real y[n];
}

parameters{
  real a;              //intercept
  real b;              //slope
  real<lower=0> sigma; //error
}

model{
  // Priors
  real mu[n];
  a ~ normal(0,10);
  b ~ normal(0,10);
  sigma ~ normal(0,10);
  // Likelihood
  for(i in 1:n){
    mu[i] = a + b * x[i];
    y[i] ~ normal(mu[i], sigma); 
  }
}
'

stan_model.1 = stan_model(model_code = stan_code.1)

stan_code.2 = '
data{
  int n; // comments
  real x[n];
  real y[n];
}

parameters{
  real a;              //intercept
  real b;              //slope
  real<lower=0> sigma; //error
}

model{
  // Priors
  real mu[n];
  a ~ normal(0,10);
  b ~ normal(0,1);
  sigma ~ normal(0,10);
  // Likelihood
  for(i in 1:n){
    mu[i] = a + b * x[i];
    y[i] ~ normal(mu[i], sigma); 
  }
}
'

stan_model.2 = stan_model(model_code = stan_code.2)

stan_code.3 = '
data{
  int n; // comments
  real x[n];
  real y[n];
}

parameters{
  real a;              //intercept
  real b;              //slope
  real<lower=0> sigma; //error
}

model{
  // Priors
  real mu[n];
  a ~ normal(0,10);
  b ~ normal(0,.1);
  sigma ~ normal(0,10);
  // Likelihood
  for(i in 1:n){
    mu[i] = a + b * x[i];
    y[i] ~ normal(mu[i], sigma); 
  }
}
'

stan_model.3 = stan_model(model_code = stan_code.3)
data = list(n = nrow(df),
            x = df$x,
            y = df$y)

fit.1 = sampling(stan_model.1, 
               data = data)
fit.2 = sampling(stan_model.2, 
               data = data)
fit.3 = sampling(stan_model.3, 
               data = data)

plot(fit.1)
plot(fit.2)
plot(fit.3)
```


## Exploring the outputs

```{r}
# Model 
stan_code = '
data{
  int n; // comments
  real x[n];
  real y[n];
}

parameters{
  real a;              //intercept
  real b;              //slope
  real<lower=0> sigma; //error
}

model{
  // Priors
  real mu[n];
  a ~ normal(0,10);
  b ~ normal(0,10);
  sigma ~ normal(0,10);
  // Likelihood
  for(i in 1:n){
    mu[i] = a + b * x[i];
    y[i] ~ normal(mu[i], sigma); 
  }
}
'

stan_model = stan_model(model_code = stan_code)

data = list(n = nrow(df),
            x = df$x,
            y = df$y)

fit = sampling(stan_model, 
               data = data)


posterior = as.matrix(fit)

str(posterior)
```

### How data explain the data

```{r}
plot(df)
abline(a = mean(posterior[,1]), b = mean(posterior[,2]))
```

```{r}
x.pred = seq(1,100)
y.cred = matrix(NA,nrow(posterior), ncol = length(x.pred))
for (i in 1:50) {
  y.cred[i, ] = posterior[i,1] + posterior[i,2] * x.pred  
}

```

Prediction intervalles

```{r}
y.pred = matrix(NA,nrow(posterior), ncol = length(x.pred))
for (i in 1:50) {
  y.pred[i, ] = rnorm(n = length(x.pred),
                      mean = y.cred[i,],
                      sd = rep(posterior[i,3], length(x.pred)))  
}


```

#### Practical

On quadratic regression


```{r}
library(rstan)
library(coda)
library(BayesianTools)

n = 100

a = 1
b = 2
c = 3
sigma = .5

x = runif(n, min = -1, max = 1)
y = a + b * x + c * x * x + rnorm(n, 0, sd = sigma)

df.quad = data.frame(
  x = x, 
  y = y
)

plot(df.quad)

# Model 
stan_code = '
data{
  int n; // comments
  real x[n];
  real y[n];
}

parameters{
  real a;              //intercept
  real b;              //slope
  real c;              //slope x2
  real<lower=0> sigma; //error
}

model{
  // Priors
  real mu[n];
  a ~ normal(0,10);
  b ~ normal(0,10);
  c ~ normal(0,10);
  sigma ~ normal(0,10);
  // Likelihood
  for(i in 1:n){
    mu[i] = a + b * x[i] + c * x[i] * x[i];
    y[i] ~ normal(mu[i], sigma); 
  }
}
'

stan_model = stan_model(model_code = stan_code)

data = list(n = nrow(df.quad),
            x = df.quad$x,
            y = df.quad$y)

fit = sampling(stan_model, 
               data = data)


print(fit)
correlationPlot(as.matrix(fit)[,1:3])

posterior = as.matrix(fit)

x.pred = df.quad$x
y.cred = matrix(NA,nrow(posterior), ncol = length(x.pred))
for (i in 1:50) {
  y.cred[i, ] = posterior[i,1] + posterior[i,2] * x.pred + posterior[i,3] * x.pred * x.pred  
}

y.pred = matrix(NA,nrow(posterior), ncol = length(x.pred))
for (i in 1:50) {
  y.pred[i, ] = rnorm(n = length(x.pred),
                      mean = y.cred[i,],
                      sd = rep(posterior[i,3], length(x.pred)))  
}

df.quad.cred = data.frame(
  x.pred,
  mean.cred = apply(y.cred, 2, function(x) mean(x, na.rm = T)),
  cred.95 = apply(y.cred, 2, function(x) quantile(x, .95, na.rm = T)),
  cred.05 = apply(y.cred, 2, function(x) quantile(x, .05, na.rm = T))
)

df.quad.pred = data.frame(
  x.pred,
  mean.pred = apply(y.pred, 2, function(x) mean(x, na.rm = T)),
  pred.95 = apply(y.pred, 2, function(x) quantile(x, .95, na.rm = T)),
  pred.05 = apply(y.pred, 2, function(x) quantile(x, .05, na.rm = T))
)
library(ggplot2)

plot(df.quad)
ggplot(data = df.quad, aes(x=x,y=y)) + 
  geom_point(data = df.quad, aes(x=x,y=y)) + 
  geom_line(data = data.frame(x = x.pred, y = mean.cred), aes(x=x,y=y), color = 'red') +
  geom_line(data = df.quad.cred, aes(x = x.pred, y = cred.95), color = 'red', lty = 2) + 
  geom_line(data = df.quad.cred, aes(x = x.pred, y = cred.05), color = 'red', lty = 2) + 
  geom_line(data = df.quad.pred, aes(x = x.pred, y = pred.95), color = 'green', lty = 2) + 
  geom_line(data = df.quad.pred, aes(x = x.pred, y = pred.05), color = 'green', lty = 2) + 
  theme_bw()

plot(df.quad$y ~ mean.cred)

```
